---
title: "Report"
editor: visual
bibliography: references.bib
---

```{r}
library(tidyverse)
library(here)
library(haven)



```

# 1. Introduction

Populist parties are gaining support around the globe and Germany is no exception. The Alternative für Deutschland (AfD) as a populist party, has expirienced a substantial increase in public support. This development is a serious problem for the german democracy. This raises questions about the contributing factors.

A commonly discussed explanation is the role of economic conditions. Economic uncertainty, declining growth and unemployment are often cited factors. This project investigates if such a relationship is measurable, with GDP, economic uncertainty and unemployment.

------------------------------------------------------------------------

# 2. Research question

> To what extent do economic fluctuations (measured by GDP, growth, or a decline in GDP) influence voting intention and the polling numbers of the AfD?

------------------------------------------------------------------------

# 3. Exploratory data analysis

```{r}
elect_clean %>% ggplot(aes(x = date, y = afd)) +
  geom_path()+
  labs(title = "AfD support in Germany")
```

This plot shows the vote share of the AfD in Germany over the years.

```{r}
unemp |> 
  ggplot(aes(unemployment_rate))+
  geom_histogram(binwidth = 0.5, fill = "forestgreen", color = "black")+
  labs(title = "Histogramm of enemployment in percent in Germany 2007 - 2026")+
  theme_minimal()
```

```{r}
allbus_mini_clean %>%
  group_by(ep01) %>%
  summarise("Total amount" = n()) |>
  rename("subjective state of the economy \n 1 = good 5 = bad" = ep01)
```
